Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-09-06 Manuscript/SS_GAM.tex
Manuscript revision

\corres{*Corresponding author name, This is sample corresponding
address. \email{xxx@uab}}

@@

\abstract{Spike-and-slab priors have been extensively studied for
variable selection, and recently been extended to function selection in
generalized additive models (GAMs). Most of the previous works on
fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC)
algorithms to identify the promising subsets of functions. The
computational burden of MCMC makes these models less feasible,
especially in high-dimensional setting. Besides, most of the current
spike-and-slab GAM consider an ``all-in-all-out'' approach for function
selection, rendering less flexibility to answer if non-linear effects
are necessary in the model. We propose Bayesian hierarchical generalized
additive models for function selection and prediction. The proposed
models employ spike-and-slab priors, mixture normal and mixture double
exponential, for smooth and sparse solution. Two EM based algorithms,
EM-IWLS and EM-Coordinate Descent, are developed for accelerated
computation. Simulation studies and real data analysis demonstrated
improved performance against the state-of-art models, mgcv and COSSO.
The software implementation of the proposed methods is freely available
via the R package BHAM.}

@@

Using additive models to interpolate nonlinear effects has a long
history. Early work of additive models can date back to 1979 when
Cleveland \citep{Cleveland1979} proposed local smoothing for bivariate
analysis. Since then, many mathematical and statistical models have been
proposed. \citep{Wood2020} With the recent advancement of computational
technology, many machine learning algorithms that are computationally
intensive start to gain much popularity in prognostic modeling and
functional selection, especially when complex signals, for example
nonlinear effects, are assumed. Such models generally have improved
prediction accuracy over classic statistical models. However, these
models have a fatal flaw, the lack of transparency and interpretability,
and hence are referred to as ``black box'' models \citep{Breiman2001}.
Given that the primary objective of biomedical studies is to understand
the etiology of diseases, ``Black box'' models are less helpful as they
can hardly explain how the change of biomarkers is associated with the
disease.

Semi-parametric regression models are a more appropriate alternative to
the ``black-box'' models, thanks to their balancing between model
flexibility and interpretability. Among those, generalized additive
models (GAMs), proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be expressed as the summation of smooth
functions, \(f_j, j = 1, \dots, p\), of a given \(p\)-dimensional vector
of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. The smooth functions can take many forms and are estimated
using a pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classic GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models assume sparsity among the
predictors, i.e.~most predictors have small or no effects on the
outcome. Further development focuses on addressing the grouping
structure among the predictors, where AMs and GAMs can be considered
special cases. Various penalties are imposed on the basis function
coefficients: Ravikumar et al. \citep{Ravikumar2009} extended the
grouped lasso \citep{Yuan2006} to additive models; Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to the
additive models. Recent Bayesian hierarchical models are also used in
the context of high-dimensional additive models. Various group
spike-and-slab priors combining with computationally intensive Markov
chain Monte Carlo (MCMC) algorithms \citep{Xu2015, Yang2020} are
proposed, where the application on AMs are by-products. Scheipl et al.
\citep{Scheipl2012} proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma, specifically for semi-parametric
regression using MCMC. The normal-mixture-of-inverse gamma prior is
mixture of t-distribution with an additional coefficients mixing vector,
where the vector follows a mixture of two Gaussian distribution. Bai et
al. \citep{Bai2020} was the first to apply group spike-and-slab lasso
prior to AMs using a fast optimization algorithm, and further
generalized the framework to GAMs\citep{Bai2021}. Regardless of the nice
statistical properties, a practical and pivotal question, if non-linear
effects exist, yet to be addressed by these methods. Besides Scheipl et
al. \citep{Scheipl2012}, most proposed methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for within group selection. In addition, a common
criticism received by these methods is that the basis function
coefficients tend to be overly penalized due to the sparsity assumption
directly applied on the basis function coefficients.\citep{Scheipl2013}
It would be of special interest to develop a fast yet flexible
generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization-based
algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively. Meanwhile,
to avoid the over-penalization of the basis function coefficients,
smoothness penalties are incorporated in the model via
re-parameterization. Smoothness penalties are commonly implemented in
GAMs through smoothing and penalized regression splines. They are
quadratic norms of the coefficients and allow locally adaptive penalties
on each smoothing function. For the most common cases of cubic splines,
the smoothness penalty conditioning on a smoothness penalty parameter
\(\lambda_j\) is a function of the integration of the second derivative
of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothness penalty
matrix, and \(\boldsymbol{\beta}_j\) is the basis function coefficients.
Smoothness penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. Marra and Wood
\citep{Marra2011} introduced a re-parameterization of the basis function
matrix based on the decomposition of the smoothness penalty matrix
\(\boldsymbol{S}\). The re-parameterization allows different penalties
imposed on the null and nonlinear spaces separately, and further allows
the sparsity in linear space and smoothness in nonlinear effect
interpolation. In addition, the priors of the linear and nonlinear
effects are independent, creating computational convenience and further
improving the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections,
but it substantially differs from previous spike-and-slab based GAMs,
i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting through optimization algorithms versus computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bilevel selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

The rest of the paper are arranged as the followings: in Chapter 2 we
demonstrate the Bayesian hierarchical generalized additive model, where
we provide the setup of the proposed model and describe the two
fast-fitting algorithms. In Chapter 3, we compare the proposed via monte
carlo simulation. .Application to real-world datasets demonstrates the
usefulness of the proposed methodologies in Chpater 4. Conclusion and
discussions are made in Chapter 5. \footnote{TODO: rewrite this
  sentence.}

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with the smoothing function \(f_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(f_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(f_j(x_j) = x_j\) with the smoothness penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of smoothing penalties from smoothing spline models. The basic idea
is to set up a smoothing penalty, depending on the smoothness penalty
matrix via Equation \ref{eq:smoothpen}, in the posterior density
function, or least squared and likelihood function in the penalized
framework. However, the direct integration of smoothness penalties into
sparsity penalties is not obvious. Given the smoothness penalty matrix
is most often symmetric and positive semi-definite \footnote{ we limit
  the discussion to univariate smoothing function. However, the penalty
  matrix could possibly contain multiple matrices, e.g.~tensor product
  smoother.}, it can be integrated into the design matrix
\(\boldsymbol{X}\) via a re-parameterization. The re-parameterization
follows Marra and Wood \citep{Marra2011}, in which eigen-decomposed the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\),
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smooth function. We further write the orthonormal matrix
\(\boldsymbol{U}\), containing the eigenvectors as column, as an (row)
augmentation of two sub-matrices
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\),
where \(\boldsymbol{U}^{\text{pen}}\) contains the eigenvectors (as
columns) for the non-zero eigenvalues, i.e.~the penalized space, and
\(\boldsymbol{U}^0\) contains the eigenvectors for the null space. We
multiply the basis function matrix \(\boldsymbol{X}\) with the
orthonormal matrix \(\boldsymbol{U}\) for the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) is scale-invariant
\footnote{ I know this work for tensor spline (see \citep{Wood2006}),
  but I am not sure if it can be generalized to all smoothing functions}.
For notation convenience, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the after re-parameterized basis function matrix for the \(j\)th
variable hereafter. Similarly, the spline function can be expressed in
the matrix form \[
f_j(x_j) = f_j^0(x_j) + f_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation \ref{eq:gam} be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} ({\beta_j^0}^T \boldsymbol{X}_j^0 + {\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

The re-parameterization sets up the foundation of the proposed model,
which has three-fold benefits. First of all, as previously mentioned,
the re-parameterization integrated the smoothness penalty into the
design matrix, and hence the penalty can be expressed in the posterior
density function implicitly in addition to the sparse penalty for
variable and function selection. Secondly, the eigendecomposition of the
smooth penalty isolates the linear space of a smooth function from the
non-linear space, improving the feasibility of bi-level
variable/function selection and estimation inference. The
eigendecomposition also introduced the linearly independent column
spaces of the smooth function matrix, which allows imposing independent
priors on the coefficients. This reduces the computational complexity
that multivariate priors induces, and greatly broadens the choices of
priors and further model choices. Last but not least, the scaling step
of \(\boldsymbol{X}^\text{pen}\) further simplifies the choice of
penalty priors, from column-specific penalties to a unified penalty.
This step is also necessary for the proposed spike-and-slab spline
prior. Conventionally, the spike-and-slab priors require scaling of
predictors \footnote{TODO: insert a citation.}

@@

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
non-linear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and non-linear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
&\node (linearTheta) [outCome] { $\theta_j^{0}\sim Beta(a, b)$ };\\
&\node (linearGamma) [outCome] { $\gamma_j^0 \sim Bin(1, \theta_j^0) $ };\\
&\node (linearBeta) [outCome] { $\beta_j^0 \sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};

\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
\node (penTheta) [outCome] { $\theta_{j}^\text{pen}\sim Beta(a, b)$ };\\
\node (penGamma) [outCome] { $\gamma_{j}^\text{pen}\sim Bin(1, \theta_j^\text{pen})$ };\\
\node (penBeta) [outCome] { $\beta_{jk}^\text{pen}\sim  DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};

\node (ab)[staticCompo] at ($(linearTheta)!0.5!(penTheta)$) {$(a, b)$};
%\node (PenaltyPrior) [PriorBoarder,  right = 1cm of ab] {part 2};
\node (s) [staticCompo]  at ($(linearBeta)!0.5!(penBeta)$)  {($s_0, s_1$)};

\node (Beta) [staticCompo, below = 1cm of s] {$\boldsymbol{\beta }= (\beta^0_1, \boldsymbol{\beta}^\text{pen}_1, \dots,\beta^0_j, \boldsymbol{\beta}^\text{pen}_j , \dots,\beta^0_p, \boldsymbol{\beta}^\text{pen}_p) $};
\node (Y) [outCome, below = 1cm of Beta] {$y_i \sim Expo. Fam. (g(\eta_i))$};

%Lines

>>>>

\corres{Nengjun Yi, Department of Biostatistics, University of Alabama
at Birmingham, Birmingham, USA. \email{nyi@uab.edu}}

@@

\abstract{There are many proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional
(p\textgreater\textgreater n) problem using group penalties and priors.
These models rarely smooth the nonlinear effect properly because of the
sparse regularization. It leads to excess shrinkage on coefficients and
inaccurate prediction. Besides, most of these GAMs consider an
``all-in-all-out'' approach for function selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address the shortcomings, using Markov chain Monte
Carlo (MCMC) algorithms for model fitting creates a new challenge,
scalability. Hence, we propose Bayesian hierarchical generalized
additive models as a solution. For smooth and sparse solutions, the
proposed models employ bi-part spike-and-slab spline priors, including
mixture double exponential and mixture normal priors. Two
optimization-based algorithms, EM-Coordinate Descent and EM-Iterative
Weighted Least Squares, are developed for expedited computation.
Simulation studies and metabolomics data analyses demonstrate improved
performance against state-of-art models, mgcv, COSSO and sparse Bayesian
GAM. The software implementation of the proposed models is freely
available via an R package BHAM.}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modelling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are not particularly attractive in biomedical studies, as they lack the
transparency to understand the etiology of diseases. In addition, how to
generalized these models to the high-dimensional setting remains
unclear.

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be expressed as the summation of smoothing
functions, \(B_j, j = 1, \dots, p\), of a given \(p\)-dimensional vector
of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}\) is the inverse of a monotonic link function. The
smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from regularized linear models, are used to fit GAMs by
accounting for the group structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models; Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to the
additive models. Recent Bayesian hierarchical models are also used in
the context of high-dimensional additive models. Various group
spike-and-slab priors combining with computationally intensive Markov
chain Monte Carlo (MCMC) algorithms \citep{Xu2015, Yang2020} are
proposed, where the application on AMs are by-products. Bai et al.
\citep{Bai2020} was the first to apply group spike-and-slab lasso prior
to Gaussian additive models(AMs) using a fast optimization algorithm,
and further generalized the framework to GAMs\citep{Bai2021}. As these
methods focus on addressing the sparsity, it is very likely to overly
penalize the basis function coefficients and produce inaccurate
predictions and curve interpolation, particularly when complex signals
are assumed and large number of knots are used. \citep{Scheipl2013} In
addition, the proposed methods adapt an `all-in-all-out' strategy,
i.e.~either including or excluding the variable completely, rendering no
space for bi-level selection. Scheipl et al. \citep{Scheipl2012}
proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothness penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothness penalties are commonly
implemented in GAMs through smoothing regression splines. They are
quadratic norms of the coefficients and allow locally adaptive penalties
on each smoothing function. The smoothness penalty conditioning on a
smoothness penalty parameter \(\lambda_j\) is a function of the
integration of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothness penalty
matrix, and \(\boldsymbol{\beta}_j\) is the basis function coefficients.
Smoothness penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of smoothing functions, also know as
the null space, from the nonlinear space. We then impose a new bi-part
spike-and-slab spline prior on the smoothing functions for bi-level
selection such that the linear and nonlinear spaces of smoothing
functions can be selected separately. The prior setup encourages a
flexible solution, rending one of three possibilities: no effect, only
linear effect, or nonlinear effect of a predictor. In addition, two fast
computing optimization-based algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies.
Analyses of two metabolomics datasets using the proposed model are
presented in Section \ref{sec:real_data}. Conclusion and discussions are
given in Section \ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothness penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of smoothing penalties from smoothing spline models. The basic idea
is to set up a smoothing penalty, depending on the smoothness penalty
matrix via Equation \ref{eq:smoothpen}, in the posterior density
function, or least squared and likelihood function in the penalized
framework. However, the direct integration of smoothness penalties into
sparsity penalties is not obvious. Given the smoothness penalty matrix
is most often symmetric and positive semi-definite \footnote{ we limit
  the discussion to univariate smoothing function. However, the penalty
  matrix could possibly contain multiple matrices, e.g.~tensor product
  smoother.}, it can be integrated into the design matrix
\(\boldsymbol{X}\) via a re-parameterization. The re-parameterization
follows Marra and Wood \citep{Marra2011}, in which eigen-decomposed the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\),
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smooth function. We further write the orthonormal matrix
\(\boldsymbol{U}\), containing the eigenvectors as column, as an (row)
augmentation of two sub-matrices
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\),
where \(\boldsymbol{U}^{\text{pen}}\) contains the eigenvectors (as
columns) for the non-zero eigenvalues, i.e.~the penalized space, and
\(\boldsymbol{U}^0\) contains the eigenvectors for the null space. We
multiply the basis function matrix \(\boldsymbol{X}\) with the
orthonormal matrix \(\boldsymbol{U}\) for the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) is scale-invariant
\footnote{ I know this work for tensor spline (see \citep{Wood2006}),
  but I am not sure if it can be generalized to all smoothing functions}.
For notation convenience, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the after re-parameterized basis function matrix for the \(j\)th
variable hereafter. Similarly, a spline function can be expressed in the
matrix form \[
f_j(x_j) = f_j^0(x_j) + f_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} ({\beta_j^0}^T \boldsymbol{X}_j^0 + {\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

The re-parameterization sets up the foundation of the proposed model,
which has three-fold benefits. First of all, as previously mentioned,
the re-parameterization integrated the smoothness penalty into the
design matrix, and hence the penalty can be expressed in the posterior
density function implicitly in addition to the sparse penalty for
variable and function selection. Secondly, the eigendecomposition of the
smooth penalty isolates the linear space of a smooth function from the
nonlinear space, improving the feasibility of bi-level variable/function
selection and estimation inference. The eigendecomposition also
introduced the linearly independent column spaces of the smooth function
matrix, which allows imposing independent priors on the coefficients.
This reduces the computational complexity that multivariate priors
induces, and greatly broadens the choices of priors and further model
choices. Last but not least, the scaling step of
\(\boldsymbol{X}^\text{pen}\) further simplifies the choice of penalty
priors, from column-specific penalties to a unified penalty. This step
is also necessary for the proposed spike-and-slab spline prior.
Conventionally, the spike-and-slab priors require scaling of predictors
\footnote{TODO: insert a citation.}

@@

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
nonlinear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and nonlinear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

@@

\matrix (linearPrior) [matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (linearGamma) [outCome] { $\gamma_j^0 \sim Bin(1, \theta_j) $ };\\
  \node (linearBeta) [outCome] { $\beta_j^0 \sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};
\matrix (penPrior) [right = 2cm of linearPrior, matrix of nodes, column sep = 0mm, row sep = 0.7cm] {
  \node (penGamma) [outCome] { $\gamma_{j}^\text{pen}\sim Bin(1, \theta_j)$ };\\
  \node (penBeta) [outCome] { $\beta_{jk}^\text{pen}\sim  DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)$};\\
};

\node (s) [staticCompo]  at ($(linearBeta)!0.5!(penBeta)$)  {($s_0, s_1$)};
\node (Beta) [staticCompo, below = 1cm of s] {$\boldsymbol{\beta }= (\beta^0_1, \boldsymbol{\beta}^\text{pen}_1, \dots,\beta^0_j, \boldsymbol{\beta}^\text{pen}_j , \dots,\beta^0_p, \boldsymbol{\beta}^\text{pen}_p) $};
\node (Theta)[outCome, above = 2cm of s] {$\theta_{j} \sim Beta(a, b)$};
\node (ab)[staticCompo, above = 0.5cm of Theta] {$(a, b)$};
\node (Y) [outCome, below = 1cm of Beta] {$y_i \sim Expo. Fam. (g(\eta_i))$};
